\section{Minimax Rule}

The Baye's rule is to minimize the average risk, while the Minimax rule is to minimize the worse case risk. 

In minimax estimation, we collapse our risk function by looking at the worse-case risk. Given $X \sim P_{\theta}$, where $\theta \in \Omega$ , and a loss function $L(\theta, d)$, we want to find an estimator $\delta$ that
minimizes the maximum risk: $\underset{\theta \in \Omega}{Sup} R(\delta, \theta)$.


\subsection{Use Baye's estimator to find Minimax rule}

The least favorable prior 

\begin{Definition}
If exists a prior $\Lambda'$, such that the Baye's risk $r_{\Lambda'} \geq r_{\Lambda}$ for any other prior distribution, then $\Lambda'$ is the least favorable prior.
\end{Definition}

\begin{Theorem}
Suppose $\delta_{\Lambda}$ is Bayes for $\Lambda$ with 

\begin{align*}
r_{\Lambda \in \Omega} &= \underset{\theta}{Sup} R(\delta_{\Lambda}, \theta)
\end{align*}

That is, the Baye's risk $r_{\Lambda}$ is the maximum risk of $\delta_{\Lambda}$. Then

1. $\delta_{\Lambda}$ is minimax rule;

2. $\Lambda$ is least favorable prior;

3. If $\delta_{\Lambda}$ is the unique Bayes estimator for $\Lambda, a. s. P_{\theta}$, then it is the unique minimax estimator. 
\end{Theorem}


The proof needs to understand the relationship of the Bayes risk and minimax risk.

\begin{align*}
 Sup R(\theta, \delta) & \geq \int R(\theta, \delta) d \Lambda(\theta) \geq \int R(\theta, \delta_{\Lambda}) d \Lambda(\theta) = \underset{\theta}{Sup} R(\delta_{\Lambda}, \theta)
\end{align*}

So we have the definition of Minimax rule
\begin{align*}
Sup R(\theta, \delta) & \geq  \underset{\theta}{Sup} R(\delta_{\Lambda}, \theta)
\end{align*}

where the first step holds because the worst-case risk of $\delta$ is greater than (or equal to) the average risk of $\delta$. 

the second step holds because $\delta_{\Lambda}$ is Bayes (and hence has an average risk no higher than that of $\delta$), 

and the third step holds because of our assumption that the Bayes risk of $\delta_{\Lambda}$ is equal to the worst-case risk. This implies that $\delta_{\Lambda}$ is minimax.

If $\delta_{\Lambda}$ is the unique Bayes estimator, then the second inequality above is strict for $\delta \neq \delta_{\Lambda}$, which implies that $\delta_{\Lambda}$ is the unique minimax.

Let $\Lambda'$ be any other prior distribution. Then we have 

\begin{align*}
r_{\Lambda'} &= \inf \int L(\theta, \delta) d \Lambda' \leq \int L(\theta, \delta_{\Lambda}) d \Lambda' \leq \underset{\theta}{Sup} R(\theta, \delta_{\Lambda}) = r_{\Lambda}
\end{align*}

The first step first and second steps are by the definition of Bayes risk, and the third step holds because the worst-case risk of $\delta_{\Lambda}$ is no less than its average risk over the distribution $\Lambda'$. Since the worst-case risk of $\delta_{\Lambda}$ is its Bayes risk over $\Lambda$ (by our assumption), we can infer that that $\Lambda$  is a least favorable prior distribution.

\begin{Corollary}
If a Bayes estimator $\delta_{\Lambda}$ has constant risk, $R(\theta, \delta_{\Lambda}) = R(\theta', \delta_{\Lambda}) $ for all $\theta, \theta'$, then $\delta_{\Lambda}$ is minimax. Note that this is a sufficient but not necessary condition.
\end{Corollary}

It is often relatively easy to check whether an estimator has constant risk, and this is typically our first line of attack for determining whether an estimator is minimax.

\subsection{RISK FUNCTION}


\begin{Example}
Let $X \sim b(n,p), 0 \leq p \leq 1$. We seek a minimax estimator of p of the form $\alpha X + \beta$, using the squared-error loss function.

Or the question could be, is the sample proportion $\frac{X}{n}$ minimax? 

We can get the risk function of sample proportion $\frac{X}{n}$, 

\begin{align*}
R(\theta, \delta) &= \frac{\theta (1- \theta)}{n}
\end{align*}

But the risk function is not constant, and reaches the maximum at $\frac{1}{2}$, while not $\frac{X}{n}$. So we could see that $\frac{X}{n}$ is not minimax.

To find the minimax, we need to assume a prior function Beta function, $B(a,b)$. The Baye's estimate under squared error loss is $\frac{X + a}{n+ a +b}$.

The risk function is only related to the distribution of $X| \theta$, while not the posterior distribution. Or we can consider that $\pi(p) = 1$. 
 \begin{align*}
R(\delta, p) &= E_{p} \Big[ (\alpha X + \beta - p)^2 \Big] = E_{p} \Big[ \alpha (X - np) + \beta + (\alpha n-1) p \Big]^2 \\
&= E_{p} \Big[ \alpha^2 (X - np)^2 + [\beta + (\alpha n-1) p]^2 + 2 \alpha (X- np) [\beta + (\alpha n -1)p] \Big] \\
&= \alpha^2 E[(X-np)^2 ] + [\beta + (\alpha n -1) p]^2 \\
&= \Big[ (\alpha n -1)^2 - \alpha^2 n \Big] p^2 + \Big[ \alpha^2 n + 2 \beta(\alpha n -1) \Big] p + \beta^2
\end{align*}

To find $\alpha, \beta$ such that $R(p, \delta)$ is constant for all $p \in H$, we see the coefficients of $p^2, p$ equal to 0 to get

 \begin{align*}
(\alpha n -1)^2 - \alpha^2 n & = 0 \\
\alpha^2 n + 2 \beta(\alpha n -1) &= 0
\end{align*}

\end{Example}

\subsection{Admissibility of Minimax Estimators}

So far we know the admissibility of Minimax estimators when the risk function is constant. Note that minimaxity does not guarantee admissibility; it only ensures the worst case
risk is optimal. We need to check for admissibility. 

\begin{Example}
Let $X_1,.. X_n$ be i.i.d $N \sim (\mu, \sigma^2)$ where $\sigma^2$ is known, and $\theta$ is the estimand. Then the minimax estimator is $\bar{X} $ under squared error loss, and we would like to determine
whether $\bar{X} $ is admissible.

We would like to ask a more general question, when is $a\bar{X} + b$ admissible?

We need to consider the prior distribution of $\theta$, generally it is Gaussian distribution. However, $\bar{X}$ is not Bayes for any prior, because under squared-error loss unbiased estimators are Bayes estimators only in the degenerate situations of zero risk, and $\bar{X}$ is unbiased.

We might try to consider the wider class of estimators $\delta_{a, \mu_0} = a \bar{X} + (1-a) \mu_0 = a\bar{X} + b$, because many of the Bayes estimators we've encountered are convex combinations of a prior and a data mean. Note however that the worst case risk for these estimators is infinite:

\begin{align*}
\underset{\theta}{Sup} E_{\theta} [(\theta - \delta(x))]^2 & =  \underset{\theta}{Sup} E_{\theta} [(\theta -\bar{X}  + \bar{X} - \delta(x))]^2 \\
& =  \underset{\theta}{Sup} E_{\theta} \Big [(\theta -\bar{X} )^2 + (\bar{X} - \delta(x))^2 \Big ]\\
& =  \underset{\theta}{Sup} E_{\theta} \Big [\frac{Var(X)}{n} + (\bar{X} - \delta(x))^2 \Big ]\\
\end{align*}


\begin{align*}
\underset{\theta}{Sup} E_{\theta} [(\theta -  \delta_{a, \mu_0})]^2 & =  \underset{\theta}{Sup} E_{\theta} [a\theta - a \bar{X} + (1-a) \theta - (1-a) \mu_0]^2 \\
& =  \underset{\theta}{Sup} E_{\theta} \Big [a^2 (\theta -\bar{X} )^2 + (1-a)^2 (\theta - \mu_0)^2 \Big ]\\
& =  \underset{\theta}{Sup} E_{\theta} \Big [a^2 \frac{Var(X)}{n} + (1-a)^2 (\theta - \mu_0)^2 \Big ]\\
&= a^2 \frac{Var(X)}{n} + (1-a)^2 \underset{\theta}{Sup} (\theta - \mu_0)^2 \\
&= + \infty
\end{align*}

\begin{itemize}
\item[(a)] 0< a < 1.

For prior distribution such as $\theta \sim N(0, m^2)$. we could prove the risk converge to a constant, $r_{\Lambda_m} \rightarrow \frac{\sigma^2}{n} =: r$. Using the conjugate prior is a good starting point,
so we let  $\Lambda_m$  be the conjugate priors with variance tending to $\infty$, so that $\Lambda_m$ tends to the (improper with $\pi(\theta) = 1$) uniform prior on R. 

The posterior for $\theta$ associated with each $\Lambda_m$,

We could see that $\delta = \frac{\frac{n \bar{X}}{\sigma^2}}{\frac{n \bar{X}}{\sigma^2} + \frac{1}{m^2}}$, which is $0 < a < 1$.

In particular, the posterior variance does not depend on $X_1,…X_n $, so Lemma 1 below automatically yields the Bayes risk

\begin{align*}
r_{\Lambda_m} &= \frac{\frac{1}{\frac{n \bar{X}}{\sigma^2} + \frac{1}{m^2}} \xrightarrow{m \rightarrow \infty} \frac{\sigma^2}{n} = \underset{\theta}{Sup} R(\theta, \bar{X}). 
\end{align*}

It follows from Theorem 1 that $\bar{X}$ is minimax and $\Lambda_m$  is least favorable.


Since we are using squared error loss, which is strictly convex, this Bayes estimator is unique. So, by Theorem 5.2.4 (which basically tells us that a unique Bayes estimator will always be admissible), $a \bar{X} + b$ is admissible.

\item[(b)] a = 0.

 In this case b is also a unique Bayes estimator with respect to a degenerate prior distribution with unit mass at $\theta = b$. So by Theorem 5.2.4, b is admissible

\item[(c)] a = 1 and b=0.

$\delta = \bar{X}$ is a minimax when prior is 1.

\item[(d)] $ a = 1, b \neq 0$.

In this case $\bar{X} + b$ is not admissible because it is dominated by $\bar{X}$. To see this, note that  $\bar{X}$ has the same variance as  $\bar{X} + b$, but strictly smaller bias.

In general, the risk of $a \bar{X} + b$
\begin{align*}
E[(a \bar{X} + b - \theta)^2] &= E [a (\bar{X}- \theta) + b + \theta(a -1)]^2 \\
&= \frac{a^2 \sigma^2}{n} + (b + \theta(a -1))^2
\end{align*}

\item[(e)] a > 1.

If we apply the above result,
\begin{align*}
E[(a \bar{X} + b - \theta)^2] &= \frac{a^2 \sigma^2}{n} + (b + \theta(a -1))^2 > \frac{\sigma^2}{n} = R(\bar{X}, \theta)
\end{align*}

The first inequality follows because the second summand in the expression for the general risk is always nonnegative. $\bar{X}$ dominates $a \bar{X} + b$  when a > 1, and so in this case  $a \bar{X} + b$  is
inadmissible.

\begin{align*}
E[(a \bar{X} + b - \theta)^2] & > (b + \theta(a -1))^2 \\
&> (a-1)^2 (\theta + \frac{b}{a-1})^2
\end{align*}

We can see that $(\theta + \frac{b}{a-1})^2$ is also a risk function, we can set $\delta = -\frac{b}{a-1}$. So, $\delta = -\frac{b}{a-1}$ dominates $a \bar{X} + b$ ,
and therefore, $a \bar{X} + b$  is again inadmissible.

\end{itemize}

\end{Example}
